---
title: "Gapminder"
author: "Michaela Fox"
date: 2025-02-26
output:
  github_document:
    toc: true
prerequisites:
  - e-vis03-boxplots
---

*Purpose*: Learning to do EDA well takes practice! In this challenge you'll further practice EDA by first completing a guided exploration, then by conducting your own investigation. This challenge will also give you a chance to use the wide variety of visual tools we've been learning.

<!-- include-rubric -->

# Grading Rubric

<!-- -------------------------------------------------- -->

Unlike exercises, **challenges will be graded**. The following rubrics define how you will be graded, both on an individual and team basis.

## Individual

<!-- ------------------------- -->

| Category | Needs Improvement | Satisfactory |
|------------------------|------------------------|------------------------|
| Effort | Some task **q**'s left unattempted | All task **q**'s attempted |
| Observed | Did not document observations, or observations incorrect | Documented correct observations based on analysis |
| Supported | Some observations not clearly supported by analysis | All observations clearly supported by analysis (table, graph, etc.) |
| Assessed | Observations include claims not supported by the data, or reflect a level of certainty not warranted by the data | Observations are appropriately qualified by the quality & relevance of the data and (in)conclusiveness of the support |
| Specified | Uses the phrase "more data are necessary" without clarification | Any statement that "more data are necessary" specifies which *specific* data are needed to answer what *specific* question |
| Code Styled | Violations of the [style guide](https://style.tidyverse.org/) hinder readability | Code sufficiently close to the [style guide](https://style.tidyverse.org/) |

## Submission

<!-- ------------------------- -->

Make sure to commit both the challenge report (`report.md` file) and supporting files (`report_files/` folder) when you are done! Then submit a link to Canvas. **Your Challenge submission is not complete without all files uploaded to GitHub.**

```{r setup}
library(tidyverse)
library(gapminder)
```

*Background*: [Gapminder](https://www.gapminder.org/about-gapminder/) is an independent organization that seeks to educate people about the state of the world. They seek to counteract the worldview constructed by a hype-driven media cycle, and promote a "fact-based worldview" by focusing on data. The dataset we'll study in this challenge is from Gapminder.

# Guided EDA

<!-- -------------------------------------------------- -->

First, we'll go through a round of *guided EDA*. Try to pay attention to the high-level process we're going through---after this guided round you'll be responsible for doing another cycle of EDA on your own!

### **q0** Perform your "first checks" on the dataset. What variables are in this

### dataset?

```{r q0-task}
## TASK: Do your "first checks" here!
?gapminder
head(gapminder)
gapminder %>%
  pull(country) %>%
  unique()
gapminder %>%
  pull(continent) %>%
  unique()
```

**Observations**:

-   Country, continent, year, lifeExp, pop, gdpPercap

### **q1** Determine the most and least recent years in the `gapminder` dataset.

*Hint*: Use the `pull()` function to get a vector out of a tibble. (Rather than the `$` notation of base R.)

```{r q1-task}
## TASK: Find the largest and smallest values of `year` in `gapminder`
year_max <- 
  gapminder %>%
  pull(year) %>%
  max()
  
year_min <- 
  gapminder %>%
  pull(year) %>%
  min()
```

Use the following test to check your work.

```{r q1-tests}
## NOTE: No need to change this
assertthat::assert_that(year_max %% 7 == 5)
assertthat::assert_that(year_max %% 3 == 0)
assertthat::assert_that(year_min %% 7 == 6)
assertthat::assert_that(year_min %% 3 == 2)

if (is_tibble(year_max)) {
  print("year_max is a tibble; try using `pull()` to get a vector")
  assertthat::assert_that(False)
}

print("Nice!")
```

### **q2** Filter on years matching `year_min`, and make a plot of the GDP per capita against continent. Choose an appropriate `geom_` to visualize the data. What observations can you make?

You may encounter difficulties in visualizing these data; if so document your challenges and attempt to produce the most informative visual you can.

```{r q2-task}
## TASK: Create a visual of gdpPercap vs continent
gapminder %>%
  filter(year == year_min) %>%
  ggplot(aes(x = continent, y = gdpPercap)) + 
  geom_boxplot() + 
  scale_y_log10()
```

**Observations**:

-   The mean for all continents falls between 1000 and 10000. The median for all continents does not pass 1\*10\^4.
-   Oceania has both the smallest GDP per capita spread and the highest median, suggesting the GDP per capita of its countries is relatively similar, or the number of countries located in Oceania is very small.
-   The Americas and Asia have noticeable outliers, with Asia having a very high outlier.
-   Asia has the largest spead.
-   All the outliers seem to be greater than the upper bound, not lower.

**Difficulties & Approaches**:

-   I originally tried a scatter plot, but the points were so close together it was difficult to tell where they were, and I was also not entirely sure were the outliers were except for one. I then tried a boxplot, which made it much easier to visualize the range of the data and see where outliers were.

### **q3** You should have found *at least* three outliers in q2 (but possibly many more!). Identify those outliers (figure out which countries they are).

```{r q3-task}
## TASK: Identify the outliers from q2
gapminder %>%
  filter(year == year_min) %>%
  filter(continent == 'Africa') %>%
  arrange(desc(gdpPercap)) %>%
  head(2)
```

```{r}
gapminder %>%
  filter(year == year_min) %>%
  filter(continent == 'Americas') %>%
  arrange(desc(gdpPercap)) %>%
  head(3)
```

```{r}
gapminder %>%
  filter(year == year_min) %>%
  filter(continent == 'Asia') %>%
  arrange(desc(gdpPercap)) %>%
  head(2)
```

```{r}
gapminder %>%
  filter(year == year_min) %>%
  filter(continent == 'Europe') %>%
  arrange(desc(gdpPercap)) %>%
  head(1)
```

**Observations**:

-   Identify the outlier countries from q2
    -   Africa: South Africa, Gabon
    -   Americas: USA, Canada, Venezuela
    -   Asia: Kuwait, Bahrain
    -   Europe: Switzerland

*Hint*: For the next task, it's helpful to know a ggplot trick we'll learn in an upcoming exercise: You can use the `data` argument inside any `geom_*` to modify the data that will be plotted *by that geom only*. For instance, you can use this trick to filter a set of points to label:

```{r layer-filter}
## NOTE: No need to edit, use ideas from this in q4 below
gapminder %>%
  filter(year == max(year)) %>%

  ggplot(aes(continent, lifeExp)) +
  geom_boxplot() +
  geom_point(
    data = . %>% filter(country %in% c("United Kingdom", "Japan", "Zambia")),
    mapping = aes(color = country),
    size = 2
  )
```

### **q4** Create a plot similar to yours from q2 studying both `year_min` and `year_max`. Find a way to highlight the outliers from q3 on your plot *in a way that lets you identify which country is which*. Compare the patterns between `year_min` and `year_max`.

*Hint*: We've learned a lot of different ways to show multiple variables; think about using different aesthetics or facets.

```{r q4-task}
## TASK: Create a visual of gdpPercap vs continent
gapminder %>%
  filter(year %in% c(year_min, year_max)) %>%
  ggplot(aes(x = continent, y = gdpPercap, group = interaction(continent, factor(year)))) + 
  geom_boxplot(aes(color = factor(year))) +  # Boxplots colored by year
  geom_point(
    data = gapminder %>% filter(year %in% c(year_min, year_max), 
                                country %in% c("South Africa", "Gabon", "United States", "Canada", "Venezuela", "Kuwait", "Bahrain", "Switzerland")),
    aes(x = continent, y = gdpPercap, color = country, shape = factor(year)),  # Differentiate years
    size = 2,
    position = position_dodge(width = 0.75)  # Properly align dots with boxplots
  ) +
  labs(color = "Country / Year", shape = "Year") +
  scale_y_log10()

```

```{r}
gapminder %>%
  filter(year %in% c(year_min, year_max)) %>%
  ggplot(aes(x = continent, y = gdpPercap, group = interaction(continent, factor(year)))) +
  geom_boxplot(aes(color = factor(year))) +  # Boxplots colored by year
  geom_point(
    data = gapminder %>% 
      filter(year %in% c(year_min, year_max), 
             country %in% c("United States", "Kuwait", "Switzerland")),  # ✅ Correct filter
    aes(x = continent, y = gdpPercap, color = country, shape = factor(year)),  # Differentiate by country and year
    size = 2,
    position = position_dodge(width = 0.75) # Properly align dots with boxplots
  ) +
  scale_color_manual(
    values = c("orange", "black", "red", "green", "blue"), # Assign colors
    labels = c(year_min, year_max, "Kuwait", "Switzerland", "United States") # Ensure proper labeling
  ) +
  labs(color = "Country / Year", shape = "Year")

```

**Observations**:

-   The median and range tended to increase between 1952 and 2007 - does this dataset take inflation into account?
-   Kuwait's GDP decreased significantly between 1952 and 2007, it is the only 1952 outlier country to show this behavior.

# Your Own EDA

<!-- -------------------------------------------------- -->

Now it's your turn! We just went through guided EDA considering the GDP per capita at two time points. You can continue looking at outliers, consider different years, repeat the exercise with `lifeExp`, consider the relationship between variables, or something else entirely.

### **q5** Create *at least* three new figures below. With each figure, try to pose new questions about the data.

```{r q5-task1}
## TASK: Your first graph - Life Expectancy vs. Time
gapminder %>%
  ggplot(aes(x = continent, y = lifeExp, group = interaction(continent, factor(year)))) + 
  geom_boxplot(aes(color = factor(year)))  # Boxplots colored by year

```

-   For all continents life expectancy increased over time.
-   This could imply that life expectancy and GDP are directly related.

```{r q5-task2}
## TASK: Your second graph - Median Life Expectancy vs. GDP for all continents
gapminder %>%
  group_by(continent, year) %>%
  mutate(med_lifeExp = median(lifeExp), med_gdpPercap = median(gdpPercap)) %>%
  ggplot(aes(x = med_gdpPercap, y = med_lifeExp)) +
  geom_line(aes(color = continent), size = 0.6)
```

-   Africa, Asia, and the Americas had sharp increases in life expectancy while their GDP did not change nearly as much.
-   Europe and Oceania were the opposite and had greater changes in GDP than life expectancy.
-   For all continents life expectancy tended to increase with GDP, giving evidence to the hypothesis made in the first graph.
-   However, the increase was different for all countries, implying that GDP was not the only factor influencing life expectancy, and vice versa.

```{r q5-task3}
## TASK: Your third graph - Life Expectancy vs. Population
gapminder %>%
  group_by(continent, year) %>%
  mutate(med_lifeExp = median(lifeExp), med_pop = median(pop)) %>%
  ggplot(aes(x = med_pop, y = med_lifeExp)) +
  geom_line(aes(color = continent), size = 0.6)
```

```{r}
gapminder %>%
  filter(country == "Mexico") %>%  # ✅ Select only Mexico
  group_by(year) %>%  # ✅ Group by year (not continent)
  ggplot(aes(x = year, y = gdpPercap)) +
  geom_line(size = 0.6) +
  labs(title = "GDP per Capita in Mexico Over Time")# ✅ Line plot for Mexico
```

```{r}
americas <- gapminder %>%
  filter(year == 2007, continent == "Americas") %>% 
  nrow()
americas
low_gdp_americas <- gapminder %>%
  filter(year == 2007, continent == "Americas", gdpPercap < 10000)%>%
  nrow()
low_gdp_americas
```

```{r}
gapminder %>%
  filter(continent == "Asia") %>%
  group_by(year) %>%
  summarize(med_GDP = median(gdpPercap), .groups = "drop") %>%
  arrange(desc(med_GDP)) %>%  # Sort in descending order
  slice_head(n = 5)  # Select the top 5 years
```

```{r}
gapminder %>%
  filter(continent == "Africa") %>%
  group_by(year) %>%
  summarize(med_GDP = median(gdpPercap), .groups = "drop") %>%
  arrange(desc(med_GDP)) %>%  # Sort in descending order
  slice_head(n = 5)  # Select the top 5 years

```

```{r}
# Compute median life expectancy for the earliest year
min_year_lifeexp <- gapminder %>%
  filter(year == year_min) %>%
  group_by(continent) %>%
  summarize(min_med_lifeExp = median(lifeExp), .groups = "drop")

# Compute median life expectancy for the latest year
max_year_lifeexp <- gapminder %>%
  filter(year == year_max) %>%
  group_by(continent) %>%
  summarize(max_med_lifeExp = median(lifeExp), .groups = "drop")

# Join datasets and calculate the difference
lifeexp_change <- min_year_lifeexp %>%
  inner_join(max_year_lifeexp, by = "continent") %>%
  mutate(lifeExp_difference = max_med_lifeExp - min_med_lifeExp)
```

-   Population and life expectancy are related, when one increases, so does the other.
-   For most continents, life expectancy plateaus once it reaches a certain point.
